---
title: "Group2_Chicago_Bike_Rental"
author: "Vinay Danepalli"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

install.packages("dplyr", "geosphere", "stringr", "httr", "tidyverse", "lubridate")
install.packages("rmarkdown", "knitr", "forecast", "xts")
install.packages("factoextra")

install.packages("ggplot2")
install.packages("nnet")
install.packages("caret")
install.packages("tensorflow")
install.packages("keras")
install.packages("randomForest")
install.packages("shinydashboard")
install.packages("deepnet")

```{r}
# load required packages -----
library(stringr)
library(httr)
library(tidyverse)
library(lubridate)
library(dplyr)
library(geosphere)
library(rmarkdown)
library(knitr)
library(forecast)
library(xts)
library(cluster)   
library(factoextra)
library(ggplot2)
library(nnet)
library(caret)
library(tensorflow)
library(keras)
library(randomForest)
library(shinydashboard)
library(deepnet)
```

# download and import data ------

```{r}
# download and import 202004
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202004-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202004 <- read.csv(unz(temp, '202004-divvy-tripdata.csv'))
unlink(temp)
```


```{r}
# download and import 202005
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202005-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202005 <- read.csv(unz(temp, '202005-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# download and import 202006
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202006-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202006 <- read.csv(unz(temp, '202006-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# download and import 202007
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202007-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202007 <- read.csv(unz(temp, '202007-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# download and import 202008
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202008-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202008 <- read.csv(unz(temp, '202008-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# download and import 202009
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202009-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202009 <- read.csv(unz(temp, '202009-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# download and import 202010
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202010-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202010 <- read.csv(unz(temp, '202010-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# download and import 202011
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202011-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202011 <- read.csv(unz(temp, '202011-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# download and import 202012
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202012-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202012 <- read.csv(unz(temp, '202012-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# download and import 202101
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202101-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202101 <- read.csv(unz(temp, '202101-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# download and import 202102
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202102-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202102 <- read.csv(unz(temp, '202102-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# download and import 202103
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202103-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202103 <- read.csv(unz(temp, '202103-divvy-tripdata.csv'))
unlink(temp)
```
# ======================================================

```{r}
# download and import 202104
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202104-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202104 <- read.csv(unz(temp, '202104-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# downlaod and import 202105
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202105-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202105 <- read.csv(unz(temp, '202105-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# downlaod and import 202106
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202106-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202106 <- read.csv(unz(temp, '202106-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# downlaod and import 202107
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202107-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)
data_202107 <- read.csv(unz(temp, '202107-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# downlaod and import 202108
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202108-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202108 <- read.csv(unz(temp, '202108-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# downlaod and import 202109
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202109-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202109 <- read.csv(unz(temp, '202109-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# downlaod and import 202110
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202110-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202110 <- read.csv(unz(temp, '202110-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# downlaod and import 202111
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202111-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202111 <- read.csv(unz(temp, '202111-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# downlaod and import 202112
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202112-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202112 <- read.csv(unz(temp, '202112-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# downlaod and import 202201
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202201-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202201 <- read.csv(unz(temp, '202201-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# downlaod and import 202202
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202202-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202202 <- read.csv(unz(temp, '202202-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
# downlaod and import 202203
dataset <- httr::GET("https://divvy-tripdata.s3.amazonaws.com/202203-divvy-tripdata.zip")
temp <- tempfile()
download.file(dataset$url, temp)

data_202203 <- read.csv(unz(temp, '202203-divvy-tripdata.csv'))
unlink(temp)
```

```{r}
print(colnames(data_202004))

# dropping columns containing ride and station ids
data_202004=data_202004[,c(-1,-6,-8)]
data_202005=data_202005[,c(-1,-6,-8)]
data_202006=data_202006[,c(-1,-6,-8)]
data_202007=data_202007[,c(-1,-6,-8)]
data_202008=data_202008[,c(-1,-6,-8)]
data_202009=data_202009[,c(-1,-6,-8)]
data_202010=data_202010[,c(-1,-6,-8)]
data_202011=data_202011[,c(-1,-6,-8)]
data_202012=data_202012[,c(-1,-6,-8)]
data_202101=data_202101[,c(-1,-6,-8)]
data_202102=data_202102[,c(-1,-6,-8)]
data_202103=data_202103[,c(-1,-6,-8)]
data_202104=data_202104[,c(-1,-6,-8)]
data_202105=data_202105[,c(-1,-6,-8)]
data_202106=data_202106[,c(-1,-6,-8)]
data_202107=data_202107[,c(-1,-6,-8)]
data_202108=data_202108[,c(-1,-6,-8)]
data_202109=data_202109[,c(-1,-6,-8)]
data_202110=data_202110[,c(-1,-6,-8)]
data_202111=data_202111[,c(-1,-6,-8)]
data_202112=data_202112[,c(-1,-6,-8)]
data_202201=data_202201[,c(-1,-6,-8)]
data_202202=data_202202[,c(-1,-6,-8)]
data_202203=data_202203[,c(-1,-6,-8)]

print(colnames(data_202004))

```


```{r}
# merge files -----
bicycle_data <- bind_rows(
  data_202004, data_202005, data_202006, data_202007, data_202008, 
  data_202009, data_202010, data_202011, data_202012, data_202101, 
  data_202102, data_202103, data_202104, data_202105, data_202106, 
  data_202107, data_202108, data_202109, data_202110, data_202111, 
  data_202112, data_202201, data_202202, data_202203)

# str() to display the structure of bicycle_data
str(bicycle_data)

```


```{r}
# number of rows before filtering null and missing values
nrows_unfiltered <- nrow(bicycle_data)
print(paste("Number of rows before filtering = ", nrows_unfiltered))
```


```{r}
# Get counts of unique values for each column 
rapply(bicycle_data,function(x)length(unique(x)))
```


```{r}
# Filter for null and missing values
bicycle_data <- bicycle_data[!(is.na(bicycle_data$rideable_type) | bicycle_data$rideable_type=="" |  
    is.na(bicycle_data$started_at) | bicycle_data$started_at=="" | 
    is.na(bicycle_data$ended_at) | bicycle_data$ended_at=="" |   
    is.na(bicycle_data$start_station_name) | bicycle_data$start_station_name=="" |   
    is.na(bicycle_data$end_station_name) | bicycle_data$end_station_name=="" |   
    is.na(bicycle_data$start_lat) | bicycle_data$start_lat=="" |      
    is.na(bicycle_data$start_lng) | bicycle_data$start_lng=="" |   
    is.na(bicycle_data$end_lat) | bicycle_data$end_lat=="" | 
    is.na(bicycle_data$end_lng) | bicycle_data$end_lng=="" | 
    is.na(bicycle_data$member_casual) | bicycle_data$member_casual==""),]
```


```{r}
# numebr of rows after filtering null and missing values
nrows_filtered <- nrow(bicycle_data)
print(paste("Number of rows after filtering = ", nrows_filtered))
```


```{r}
# Number of rows deleted 
print(paste("Number of rows deleted = ", nrows_unfiltered - nrows_filtered))
```


```{r}
# summary of bicycle data
summary(bicycle_data)
```

```{r}
# Get counts of unique values for each column 
rapply(bicycle_data,function(x)length(unique(x)))
```

```{r}
# Get unique rideable type
unique(bicycle_data$rideable_type)
```


```{r}
# deriving the weekday of the ride start time from started_at column 
week_days <- weekdays(as.Date(bicycle_data$started_at))
bicycle_data <- cbind(bicycle_data, week_days)

# Add ride_duration duration column in the data frame
bicycle_data$ride_duration <- difftime(bicycle_data$ended_at, bicycle_data$started_at, units = "hours") 


# Format character start and end times into Datetime.
bicycle_data$started_at = as_datetime(bicycle_data$started_at)
bicycle_data$ended_at = as_datetime(bicycle_data$ended_at)

bicycle_data$ride_duration <- as.numeric(bicycle_data$ride_duration)

bicycle_data$start_date <- as.Date(bicycle_data$started_at, format="%Y-%m-%d")
bicycle_data$year <- year(bicycle_data$start_date)
bicycle_data$month <- month(bicycle_data$start_date)
bicycle_data$day <- day(bicycle_data$start_date)

#Adding days of the week column : weekday or weekend
bicycle_data$week <- ifelse(week_days %in% c("Saturday", "Sunday"), "weekend", "weekday")
str(bicycle_data)
```

```{r}
# Save bicycle_data as csv and load it again. 
write.csv(bicycle_data, "cleaned_bicycle_data.csv", row.names=FALSE)
bicycle_data <- read.csv("cleaned_bicycle_data.csv")
```


```{r}
ggplot(bicycle_data, aes(x=rideable_type, fill = member_casual))+
  geom_bar(stat="count", width=0.7, position = position_dodge())+
  labs(title = "#Rides vs Membership", y = "#Rides", 
      x = "Rideable type") +
  theme(axis.text.x = element_text(angle = 90), 
      plot.title = element_text(color="black", size=15, face="bold.italic"),
      axis.title.x = element_text(color="black", size=14, face="bold"),
      axis.title.y = element_text(color="black", size=14, face="bold")) +
        facet_wrap(~year)
```


```{r}
# counts of rides having  start_station_name != end_station_name
common_route <- rbind(bicycle_data %>% 
  filter(start_station_name != end_station_name, member_casual == "member" ) %>%
    group_by(start_station_name, end_station_name, member_casual, rideable_type) %>%
    summarise(n = n(), .groups = "drop") %>%
    arrange(desc(n)) %>%
    head(10), 
  bicycle_data %>% 
  filter(start_station_name != end_station_name, member_casual == "casual" ) %>%
    group_by(start_station_name, end_station_name, member_casual, rideable_type) %>%
    summarise(n = n(), .groups = "drop") %>%
    arrange(desc(n)) %>%
    head(10))
  
common_route %>%
    ggplot(aes(x = reorder(start_station_name,-n), y = n, fill = member_casual)) +
    geom_bar(stat = "identity")+
    geom_text(aes(label=n), vjust=1.6, color="black", size=1) +
    labs(title = "#Rides vs Membership", y = "#Rides", 
      x = "Check-out station != Check-in station") +
    theme(axis.text.x = element_text(angle = 90), 
        plot.title = element_text(color="black", size=15, face="bold.italic"),
        axis.title.x = element_text(color="black", size=14, face="bold"),
        axis.title.y = element_text(color="black", size=14, face="bold")) +
          facet_grid(rideable_type ~ member_casual)

common_route

```


```{r}
# counts of rides having same start_station_name and end_station_name
rbind(bicycle_data %>% 
  filter(start_station_name == end_station_name, member_casual == "member" ) %>%
    group_by(start_station_name, member_casual, rideable_type) %>%
    summarise(n = n(), .groups = "drop") %>%
    arrange(desc(n)) %>%
    head(10), 
  bicycle_data %>% 
  filter(start_station_name == end_station_name, member_casual == "casual" ) %>%
    group_by(start_station_name, member_casual, rideable_type) %>%
    summarise(n = n(), .groups = "drop") %>%
    arrange(desc(n)) %>%
    head(10)) %>%
    ggplot(aes(x = reorder(start_station_name,-n), y = n, fill = member_casual)) +
    geom_bar(stat = "identity")+
    geom_text(aes(label=n), vjust=1.6, color="black", size=1) +
    labs(title = "#Rides vs Membership", y = "#Rides", 
      x = "Check-out station == Check-in station") +
    theme(axis.text.x = element_text(angle = 90), 
        plot.title = element_text(color="black", size=15, face="bold.italic"),
        axis.title.x = element_text(color="black", size=14, face="bold"),
        axis.title.y = element_text(color="black", size=14, face="bold")) +
          facet_grid(rideable_type ~ member_casual)
    
```


```{r}
# Places where bikes are rented more than 30 days

bicycle_data %>% 
  arrange(desc(ride_duration)) %>% filter(ride_duration > 30 * 24) %>% head(10) %>%
    ggplot(aes(x = reorder(start_station_name,-ride_duration), y = ride_duration/24, fill = member_casual)) +
    geom_bar(stat = "identity")+
    labs(title = "Membership vs Rideable_type vs #Days_rented ", y = "Days rented for", 
      x = "Checked out at") +
    theme(axis.text.x = element_text(angle = 90), 
        plot.title = element_text(color="black", size=15, face="bold.italic"),
        axis.title.x = element_text(color="black", size=14, face="bold"),
        axis.title.y = element_text(color="black", size=14, face="bold")) +
          facet_wrap(rideable_type ~ member_casual)
```


```{r}
rbind(bicycle_data %>%
  filter(week =="weekend") %>%
  group_by(week, member_casual, start_station_name, end_station_name, rideable_type) %>%
  summarise(n = n(), .groups = "drop") %>% 
  arrange(desc(n)) %>% head(10),
bicycle_data %>%
  filter(week =="weekday") %>%
  group_by(week, member_casual, start_station_name, end_station_name, rideable_type) %>%
  summarise(n = n(), .groups = "drop") %>% 
  arrange(desc(n)) %>% head(10)) %>%
  ggplot(aes(x = reorder(start_station_name, -n), y = n, fill = member_casual)) +
    geom_bar(stat = "identity")+
    labs(title = "Membership vs Rideable_type vs Week", y = "#Rides", 
      x = "Checked out at") +
    theme(axis.text.x = element_text(angle = 90),
        plot.title = element_text(color="black", size=15, face="bold.italic"),
        axis.title.x = element_text(color="black", size=14, face="bold"),
        axis.title.y = element_text(color="black", size=14, face="bold")) +
          facet_grid(week ~ member_casual + rideable_type)
```


# Part 2 
```{r}
## Find the number of check-outs per station
start_station_name <- unique(divvy$start_station_name)
head(start_station_name)

# ## Find the number of check-ins per station
end_station_name <- unique(divvy$end_station_name)
head(end_station_name)
```


```{r}
str(start_station_name)
str(end_station_name)
start_station_name[!(start_station_name %in% end_station_name)]
end_station_name[!(end_station_name %in% start_station_name)]
```


```{r}
#Top 5 stations with most check outs
start_station_counts <- divvy %>% group_by(start_station_name) %>% summarise(n = n()) %>% ungroup() %>% arrange(desc(n))
head(start_station_counts, 5)
#Top 5 stations with least check outs
tail(start_station_counts, 200)
```


```{r}
#Top 5 stations with most check ins 
end_station_counts <- divvy %>% group_by(end_station_name) %>% summarise(n = n()) %>% ungroup() %>% arrange(desc(n))
head(end_station_counts, 5)
#Top 5 stations with least check ins 
tail(end_station_counts, 200)
```


```{r}
## Find the number of unique days
unique_days <- unique(divvy$start_date)
head(unique_days)
str(unique_days)
```

```{r}
#total rented duration per day
duration_per_day <- divvy %>% group_by(start_date) %>% summarise(ride_duration = sum(ride_duration))%>% ungroup()
head(duration_per_day)
class(duration_per_day)

duration_per_day <- duration_per_day %>%
            filter(ride_duration > 0)

```

```{r}
divvy_xts <- xts(duration_per_day$ride_duration, order.by = duration_per_day$start_date)
fig1 <- autoplot(divvy_xts) + ggtitle("Divvy daily ridership duration") + ylab("Hours per day") + xlab("Year")
fig1
```

```{r}
divvy_ts <- ts(divvy_xts, start = 2020, frequency = 365)
fig2 <- ggseasonplot(divvy_ts) + ggtitle("Seasonal plot of Divvy daily ridership")
fig2
```

```{r}
plot(divvy_xts['2020-05-01/2020-05-28'])
```

```{r}
divvy_w_ts <- ts(divvy_xts['2020-04-01/2021-03-31'], start = c(2020, 1), frequency = 7)
fig3 <- ggsubseriesplot(divvy_w_ts) + ggtitle("Subseries plot for weekly seasonality")
fig3
```

```{r}
divvy_w <- divvy %>% group_by(week = as.POSIXct(cut(start_date, "week"))) %>% summarise(weekly_rides = sum(ride_duration)) 
divvy_m <- divvy %>% group_by(month = as.POSIXct(cut(start_date, "month"))) %>% summarise(monthly_rides = sum(ride_duration))

divvy_d_test <- divvy_xts['2020-10-01/2021-03-31']
divvy_w_test <- divvy_w %>% filter(week >= "2020-10-01")
divvy_m_test <- divvy_m %>% filter(month >= "2020-10-01")

divvy_d_train <- divvy_xts['2020/2020-10-01']
divvy_w_train <- divvy_w %>% filter(week < "2020-10-01")
divvy_m_train <- divvy_m %>% filter(month < "2020-10-01")

divvy_m_train <- xts(divvy_m_train$monthly_rides, order.by = divvy_m_train$month)

d_ts <- ts(divvy_m_train, start = 2020, frequency = 12)
fit1 <- ets(d_ts, damped = TRUE)
summary(fit1)
```

```{r}
fig4 <- autoplot(forecast(fit1, h = 6))
fig4
```

```{r}
# ?ets
fit1b <- ets(d_ts, model = c("AAA"), damped = TRUE)
summary(fit1b)
```

```{r}
fc1 <- forecast(fit1b, h = 6)
plot(fc1, showgap = FALSE)
```

```{r}
#arima
ggtsdisplay(divvy_m_train, main="Monthly Divvy Usage")
```

```{r}
ggtsdisplay(diff(divvy_m_train))

```


```{r}
fit5 <- Arima(divvy_m_train, order = c(1,1,0), seasonal = list(order = c(0,1,0), period = 6))

fc5 <- forecast(fit5, h=3)
plot(fc5, showgap = FALSE)
```



